---
title: "Lab 4: Data Wrangling"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1- Answer questions on M3/A3
2- Answer questions on M4
3- Practice wrangling datasets with dplyr functions

## Set up your session

Today we will work with a dataset from the [North Temperate Lakes Long-Term Ecological Research Station](https://lter.limnology.wisc.edu/about/overview). The NTL-LTER is located in the boreal zone in northern Wisconsin, USA. We will use the [chemical and physical limnology dataset](https://lter.limnology.wisc.edu/content/cascade-project-north-temperate-lakes-lter-core-data-physical-and-chemical-limnology-1984), running from 1984-2016. 

Opening discussion: why might we be interested in long-term observations of temperature, oxygen, and light in lakes?

> Add notes here: 

>Look at metadata!

> Peter (experimental) and Paul (control) Lakes right next to each other. Change biology, see what else changes. Nutrient data (N, P, DO), physical data (irradiance (light penetration)). 

>Import, explore, wrangle
* Sunset for Peter and Paul Lakes
* Fix dates
* Filtering (on multiple values with %in%)

>Need to understand the system to be a good data scientist. Where are the nutrient inputs coming from?

```{r, message = FALSE}
getwd()
#install.packages(tidyverse)
library(tidyverse)
#install.packages(lubridate)
library(lubridate)
NTL.phys.data <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

class(NTL.phys.data$sampledate)
# Format sampledate as date
NTL.phys.data$sampledate <- as.Date(NTL.phys.data$sampledate, format = "%m/%d/%y")
```


## Filter

Filtering allows us to choose certain rows (observations) in our dataset.

```{r}
# determine class of what you're interested in. This will help you understand what kind of filtering clause to use. Ex. numbers & characters can use "%in%"
class(NTL.phys.data$lakeid)
class(NTL.phys.data$depth)

# dplyr filtering: "where clause"
NTL.phys.data.surface <- filter(NTL.phys.data, depth == 0)

# Choose multiple conditions to filter. "filter lakename in Paul Lake and Peter Lake"
summary(NTL.phys.data$lakename)
NTL.phys.data.PeterPaul <- filter(NTL.phys.data, lakename %in% c("Paul Lake", "Peter Lake"))

# Choose a range of conditions of a numeric or integer variable
summary(NTL.phys.data$daynum)
NTL.phys.data.JunethruOctober <- filter(NTL.phys.data, daynum %in% c(152:304))

# Exercise 1: 
# filter NTL.phys.data for the year 1999
# what code do you need to use, based on the class of the variable?
class(NTL.phys.data$year4)

NTL.phys.data.1999 <- filter(NTL.phys.data, year4 == 1999)

# Exercise 2: 
# filter NTL.phys.data for Tuesday Lake from 1990 through 1999.
class(NTL.phys.data$lakename)

NTL.phys.data.Tues.1990thru1999 <-
  filter(NTL.phys.data, year4 %in% c(1990:1999) &
                        lakename == "Tuesday Lake")
```
Question: Why don't we filter using row numbers?

> Answer: Row numbers are somewhat arbitrary and do not necessarily match our data. 

## Pipes

Pipe is another method to wrangle datasets that looks cleaner and is easier to read.  We designate a pipe with `%>%`. A good way to think about the function of a pipe is with the word "then." 

Let's say we want to take our raw dataset (NTL.phys.data), *then* filter the data for Peter and Paul lakes, *then* select temperature and observation information, and *then* add a column for temperature in Fahrenheit: 

```{r}
NTL.phys.data.processed <- 
  NTL.phys.data %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") %>%
  select(lakename, sampledate:temperature_C) %>%
  mutate(temperature_F = (temperature_C*9/5) + 32)

#Exercise 3: Using a pipe filter NTL.phys.data for Tuesday Lake from 1990 through 1999 only for July 

NTL.phys.data.Tues.1990thru1999.July <- NTL.phys.data.Tues.1990thru1999 %<%
  filter(month(sampledate))



#Exercise 4: Using the data from part 3, a pipe and the summarise() function find the mean surface temperature (hints: you will need to add another filter for depth==0). Make sure you eliminate NAs before computing the means
Exercise4 <- Exercise3 %>% 
filter(depth == 0) %>% 
  filter(!is.na(temperature_C)) %>%    #alternatively: dropna(temperature_C)
  summarise(mean = mean(temperature_C))

#=22.64
  
```

## Gather and Spread

For gather we will use `pivot_longer` (columns into rows) and for spread we will use `pivot_wider` (rows into columns). 

```{r}
#Exercise 5: gather irradiance data (measured in the water column and measured on the deck of the sampling boat) into one column using pivot_longer


#Exercise 6: spread temperatureC into more than one column based on the depth

```

